Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2015-12-17 jss/article.tex
Added examples
[skip ci]

%multisample_enrichment<-function(Genes_by_sample=NULL,
%                                 maps = ACSNMineR::ACSN_maps, 
%                                 correction_multitest = "BH",
%                                 statistical_test = "fisher",
%                                 min_module_size = 5,
%                                 universe = "map_defined",
%                                 threshold = 0.05,
%                                 cohort_threshold = TRUE,
%                                 alternative = "greater"){
It may be of interest to compare enrichment of pathways in different cohorts or experiments. For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
To facilitate such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function. It relies on the \code{enrichment} function but takes a list of character vector genes. The name of each element of the list will be assumed to be the name of the sample for further analysis.
Most of the arguments given to \code{multisample_enrichment} are the same as the ones passed to \code{enrichment}. However, the \code{cohort_threshold} is designed to filter out modules which would not pass the significance threshold in all samples.

%represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log", 
%                               low = "steelblue" , high ="white",
%                              nrow = 1,sample_name = "Sample",
%                               na.value = "grey"){

Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
allows representation of results from \code{enrichment} or
\code{multisample_enrichment} with a limited number of parameters. For
examples, two types of display are available: heat-map tiles or bars. For
multiple samples using a barplot representation the number of rows used can be
provided, all plots will be on the same row otherwise. For the heatmap, the
color of the non-significant modules, and boundaries of the gradient for
significant values can also be tuned.

The package also includes a use-case that we detail below. Using,
\code{genes_test} it is possible to explore the different possibilities of
exploration of the maps. We previously computed the p-value of
\code{genes_test} with default parameters.  The number of modules which have a
p-value below 0.05 was 36, that can be compared to the 49 obtained without
correction: \code{enrichment(genes_test,correction_multitest = FALSE)}

\begin{table}[h!]
  \centering
  \caption{Overview of results from enrichment analysis without correction.
  Module : name of the module. Mod. size: size of the module. Genes in module:
  genes from input which are found in the module. p-value: uncorrected p-value.
  Test : null hypothesis used, greater is synonym of enrichment.}
  \label{tab:table2}

  \begin{tabular}{l|c|c|c|c}
	\hline
Module & Mod. size & Genes in module & p-value & Test\\
	\hline
Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 & 1.04e-08 & greater \\
               &   &  KAT2B                 &          &         \\
Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.04e-02 & greater \\
Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.98e-03 & greater \\
	\hline
	\end{tabular}
\end{table}

@@

\begin{verbatim}
> library(RNaviCell)
> navicell <- NaviCell()
> navicell$launchBrowser()

\end{verbatim}

>>>>

%multisample_enrichment<-function(Genes_by_sample=NULL,
%                                 maps = ACSNMineR::ACSN_maps, 
%                                 correction_multitest = "BH",
%                                 statistical_test = "fisher",
%                                 min_module_size = 5,
%                                 universe = "map_defined",
%                                 threshold = 0.05,
%                                 cohort_threshold = TRUE,
%                                 alternative = "greater"){
It may be of interest to compare enrichment of pathways in different cohorts or experiments. For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
To ease such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function. It relies on the \code{enrichment} function but takes a list of character vector genes. The name of each element of the list will be assumed to be the name of the sample for further analysis.
Most of the arguments given to \code{multisample_enrichment} are the same as the ones passed to \code{enrichment}. However, the \code{cohort_threshold} is designed to filter out modules which would not pass the significance threshold in all samples.

%represent_enrichment<-function(enrichment, plot = "heatmap" , scale = "log", 
%                               low = "steelblue" , high ="white",
%                              nrow = 1,sample_name = "Sample",
%                               na.value = "grey"){
Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It allows representation of results from \code{enrichment} or \code{multisample_enrichment} with a limited number of parameters. For examples, two types of display are available: heat-map tiles or bars. For multiple samples using a barplot representation the number of rows used can be provided, all plots will be on the same row otherwise. For the heatmap, the color of the non-significant modules, and boundaries of the gradient for significant values can also be tuned.

The package also includes a use-case that we detail below. Using, \code{genes_test} it is possible to explore the different possibilities of exploration of the maps. We previously computed the p-value of \code{genes_test} with default parameters.
The number of modules which have a p-value below 0.05 was 36, that can be compared to the 49 obtained without correction:
\code{enrichment(genes_test,correction_multitest = FALSE)}
\begin{table}[h!]
  \centering
  \caption{Overview of results from enrichment analysis without correction. Module : name of the module. Module size: size of the module. Genes in module: genes from input which are found in the module. p-value: uncorrected p-value. Test : null hypothesis used, greater is synonym of enrichment.}
  \label{tab:table2}

  \begin{tabular}{l|c|c|c|c}
	\hline
Module & Module size & Genes in module & p-value & Test\\
	\hline
Apoptosis:E2F1 & 5 &  ATM E2F1 EP300 KAT2B  & 1.043478e-08 & greater \\
Apoptosis:G2 M CHECKPOINT & 35  & ATM KAT2B & 2.040546e-02 & greater \\
Apoptosis:WNT CANONICAL & 53 & E2F1 EP300 KAT2B  & 4.984433e-03 & greater \\
	\hline
	\end{tabular}
\end{table}

We can now plot the first ten rows of the results obtained for corrected and uncorrected fisher test (figure 1).

\begin{figure}
	\centering
	\caption{Representation of the enriched modules (first ten rows), either with either Bonferroni correction or no correction. Grey tiles means that the data is not available for this module in this sample. P-values of low signifance are in white, whereas p-values of high significance are represent in blue. }
	\includegraphics{comparison_corrected_unc.pdf}
\end{figure}

@@
